*flippant* -- An R package for the automated analysis of dithionite scramblase

### 2015-10-05 inst/manuscript/manuscript.rmd
Make execution(more?) reliable by introducing empyt lines;

#Availablility and Implementation
*flippant* has been implemented in R [@r_core_team_r:_2015] and integrated 
into the `bioconductor` collection of tools for biomedical data analysis [@huber_orchestrating_2015]. It is thus install- and loadable as follows from 
within a functional R environment:
```{r, eval = FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(pkgs = "flippant")
library(flippant)
```
The package depends on the following packaged extensions to R base 
functionality, which are automatically satisfied on installation: 
`r flippant_dependencies`.

@@

#Main Features
##Template Generation for Data Input
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program-compatible tab-delimited plain text template, 
including column names, commentary, expected data type and the default used when
omitted.
```{r, code=formatR::usage('scramblaseAssayInputTemplate'), eval=FALSE}
```
The application agnostic tab-delimited format must be maintained for the 
completed table, which is subjected to extensive input checks when read by 
*flippant*.

##Graphic Representation of Spectral Traces
In a first, graphical data analytic step of a scramblase assay-derived data 
set, fluorescent spectra or traces may be plotted in publication-appropriate
quality with acquisition time in the x- and fluorescence intensity on the y-axis
using
```{r, code = formatR::usage('scramblaseAssayTraces'), eval = FALSE}
```
and handing the function the path to the user assembled spreadsheet of input 
data (`x`). The `adjust` argument to the function call serves to selectively 
inactivate *wmtsa*-provided (see above) alignment of multiple spectra to the 
time of dithionite addition. Acquisition time coverage of the plot may be tuned 
by user-provided minimal (`timeMinSec` ) and maximal (`timeMaxSec`) time points
to be (in seconds) included.

##Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart 
of the analysis of dithionite scramblase assays, allowing for the visual 
comparison of specific scramblase activity between interrogated experimental 
series, which may, for example, be representative of protein preparations including 
mutant protein forms or crude extracts with or without depletion of a candidate
scramblase. 
```{r, code = formatR::usage('scramblaseAssayPlot'), eval = FALSE}
```
allows for the complete generation of the PPR plot in publication quality from 
the user assembled spreadsheet of input data (`x`), including automatically 
added graphical representations the PPR value equal to the fit constant 
$\alpha$, which implies all liposomes on average carrying one scramblase and 63%
have one or more [i.e. are active; @goren_constitutive_2014]. The behavior of the 
function may be adjusted to use scramblase-mediated fluorescence reduction 
normalized to either the maximum measurement in the series (`scaleTo = "data"`)
or the plateau of an exponential fit to it (`scaleTo = "model"`; for reasoning
see above) and whether or not to require the exponential curve fitted to the 
data to go through the origin.

Numeric scramblase assay results such as $\alpha$ may be generated in 
tabular form using
```{r, code = formatR::usage('scramblaseAssayStats'), eval = FALSE}
```
with the same modifying arguments to the function as for `scramblaseAssayPlot`.

@@

*flippant*s capabilities are demonstrated here by reanalyzing data underlying 
Figure 4C/D in @ploier_perturbed_submitted. The following work flow in a first 
step plots the corresponding spectral traces (Figure 1A). Note that algorithmic
alignment to the time of dithionite addition is employed (default) as well as 
custom trimming of the time axis to account for divergent acquisition 
times between experiments.
```{r, eval = FALSE}
# Plot spectral traces (Figure 1A)
scramblaseAssayTraces("inputTable.txt", timeMaxSec = 350)
```
```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "***flippant*-based reanalysis of data in Figure 4C/D from @ploier_perturbed_submitted.** (A) represents spectral traces as produced by a call to the function `scramblaseAssayTraces`, while (B) shows protein to phospholipid ration (PPR) plots for the data using a x-axis trimmed to a maximum value of 2 (see text for details)."}
setwd(tmpDir)
figure1 <- addpanel(
  figure = figure1,
  topPanel = 1,
  bottomPanel = 3,
  leftPanel = 1, 
  rightPanel = 3, 
  panel = scramblaseAssayTraces("inputTable.txt", timeMaxSec = 350) + theme_bw())
figure1 <- addpanel(
  figure = figure1,
  topPanel = 4,
  bottomPanel = 5,
  leftPanel = 1, 
  rightPanel = 3, 
  panel = scramblaseAssayPlot("inputTable.txt") + 
    theme_bw() + 
    xlim(0, 2) + 
    xlab(expression("PPR"~~("mg"/"mmol"))) +
    ylab(expression(P[{phantom()>=1}])))
grid.draw(figure1)
```

Scramblase activity analysis as described above is performed on the data set with
the following calls.
```{r, eval = FALSE}
# Plot PPR plot (Figure 1B)
scramblaseAssayPlot("inputTable.txt") + xlim(0, 2)
# Write out results in tabular manner
scramblaseAssayStats("inputTable.txt")
```
```{r, echo = FALSE}
setwd(tmpDir)
max_PPR <- read.delim(
    "inputTable.txt",
    check.names = FALSE) %>%
  flippant:::calculatePpr() %>%
  max(
    na.rm = TRUE) %>%
  round(
    digits = 2
  )
```
The first call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 1B. Using the customization facilities provided by `ggplot2` 
[@wickham_ggplot2_2009] as a plotting system the plot's x-axis is trimmed to
avoid visual compression of all data by a single data point from the wild type 
rhodopsin series in experiment 1, where $\text{PPR}=`r max_PPR`$. Dashed lines
are added automatically as graphical representations of the fit constant 
$\alpha$, but are difficult to evaluate visually.

Recapitulation of the calculations underlying PPR plot generation by 
`scramblaseAssayPlot` in a call to `scramblaseAssayStats` produces tabular 
output represented in Table 1, which includes series-specific fit constants 
$\alpha$ in numeric and easily accessible form.
```{r, echo = FALSE}
setwd(tmpDir)
ppr_results <- scramblaseAssayStats("inputTable.txt")
ppr_results_alpha_average <- ppr_results %>%
  split(ppr_results$`Experimental Series`) %>% 
  sapply(function(y){mean(y$`Fit Constant`, na.rm = TRUE)}) %>%
  round(2)
ppr_results_alpha_average_names <- names(ppr_results_alpha_average) %>%
  head(n = length(ppr_results_alpha_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(names(ppr_results_alpha_average), n = 1))
ppr_results_alpha_average_print <- ppr_results_alpha_average %>%
  unname() %>%
  head(n = length(ppr_results_alpha_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(ppr_results_alpha_average, n = 1))
knitr::kable(
  ppr_results,
  row.names = FALSE,
  caption = "**Tabular output of *flippant*-based reanalysis of data in Figure 4C/D from @ploier_perturbed_submitted.** An plain text-version of this output is produced by calling `scramblaseAssayStats`.")
```

@@

#Conclusion
```{r, echo = FALSE, cache = TRUE}
setwd(tmpDir)
analysis_timing_n <- 100
analysis_timing_function <- function(){
  scramblaseAssayTraces(
    "inputTable.txt",
    timeMaxSec = 350)
  scramblaseAssayPlot("inputTable.txt")
  scramblaseAssayStats("inputTable.txt")
  gc()
}
analysis_timing <- microbenchmark(
  analysis_timing_function(),
  times = analysis_timing_n)

@@

analysis_timing_iqr <- round(
  diff(
    unlist(
      summary(analysis_timing)[c("lq", "uq")],
      use.names = FALSE)),
  2)
```
With *flippant* we present an integrated solution for data analysis in the 
context of dithionite scramblase assays. Requiring only basic familiarity with 
R as an environment for statistical computing and graphics 
[@r_core_team_r:_2015], scientists can quickly analyse such data and arrive at 
publication-grade graphics that offer extensive facilities for individual 
optimization and adaptation [@wickham_ggplot2_2009]. On a laptop system with 
`r parallel::detectCores()` cores and 8 GB RAM (no SSD), running `Linux` and 
R version `r paste(R.version[c("major","minor")], collapse = ".")`, the
statements generating equivalents to the plots and calculations presented here
take a mere `r analysis_timing_median` seconds (median; 
$\text{IQR}=`r analysis_timing_iqr`~\text{s}$; $n=`r analysis_timing_n`$).

>>>>

#Availablility and Implementation
*flippant* has been implemented in R [@r_core_team_r:_2015] and integrated 
into the `bioconductor` collection of tools for biomedical data analysis [@huber_orchestrating_2015]. It is thus install- and loadable as follows from 
within a functional R environment:

```{r, eval = FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(pkgs = "flippant")
library(flippant)
```

The package depends on the following packaged extensions to R base 
functionality, which are automatically satisfied on installation: 
`r flippant_dependencies`.

@@

#Main Features
##Template Generation for Data Input
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program-compatible tab-delimited plain text template, 
including column names, commentary, expected data type and the default used when
omitted.

```{r, code=formatR::usage('scramblaseAssayInputTemplate'), eval=FALSE}
```

The application agnostic tab-delimited format must be maintained for the 
completed table, which is subjected to extensive input checks when read by 
*flippant*.

##Graphic Representation of Spectral Traces
In a first, graphical data analytic step of a scramblase assay-derived data 
set, fluorescent spectra or traces may be plotted in publication-appropriate
quality with acquisition time in the x- and fluorescence intensity on the y-axis
using

```{r, code = formatR::usage('scramblaseAssayTraces'), eval = FALSE}
```

and handing the function the path to the user assembled spreadsheet of input 
data (`x`). The `adjust` argument to the function call serves to selectively 
inactivate *wmtsa*-provided (see above) alignment of multiple spectra to the 
time of dithionite addition. Acquisition time coverage of the plot may be tuned 
by user-provided minimal (`timeMinSec` ) and maximal (`timeMaxSec`) time points
to be (in seconds) included.

##Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart 
of the analysis of dithionite scramblase assays, allowing for the visual 
comparison of specific scramblase activity between interrogated experimental 
series, which may, for example, be representative of protein preparations including 
mutant protein forms or crude extracts with or without depletion of a candidate
scramblase.

```{r, code = formatR::usage('scramblaseAssayPlot'), eval = FALSE}
```

allows for the complete generation of the PPR plot in publication quality from 
the user assembled spreadsheet of input data (`x`), including automatically 
added graphical representations the PPR value equal to the fit constant 
$\alpha$ ($P_{\geq 1} = 1 - e^{-1}$), which implies all liposomes on average 
carrying one scramblase and 63% have one or more 
[i.e. are active; @goren_constitutive_2014]. The behavior of the function may be
adjusted to use scramblase-mediated fluorescence reduction normalized to either
the maximum measurement in the series (`scaleTo = "data"`) or the plateau of an
exponential fit to it (`scaleTo = "model"`; for reasoning see above) and whether
or not to require the exponential curve fitted to the data to go through the
origin.

Numeric scramblase assay results such as $\alpha$ may be generated in 
tabular form using

```{r, code = formatR::usage('scramblaseAssayStats'), eval = FALSE}
```

with the same modifying arguments to the function as for `scramblaseAssayPlot`.

@@

*flippant*s capabilities are demonstrated here by reanalyzing data underlying 
Figure 4C/D in @ploier_perturbed_submitted. The following work flow in a first 
step plots the corresponding spectral traces (Figure 1A). Note that algorithmic
alignment to the time of dithionite addition is employed (default) as well as 
custom trimming of the time axis to account for divergent acquisition 
times between experiments.

```{r, eval = FALSE}
# Plot spectral traces (Figure 1A)
scramblaseAssayTraces("inputTable.txt", timeMaxSec = 350)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "***flippant*-based reanalysis of data in Figure 4C/D from @ploier_perturbed_submitted.** (A) represents spectral traces as produced by a call to the function `scramblaseAssayTraces`, while (B) shows protein to phospholipid ration (PPR) plots for the data using a x-axis trimmed to a maximum value of 2 (see text for details)."}
setwd(tmpDir)
figure1 <- addpanel(
  figure = figure1,
  topPanel = 1,
  bottomPanel = 3,
  leftPanel = 1, 
  rightPanel = 3, 
  panel = scramblaseAssayTraces("inputTable.txt", timeMaxSec = 350) + theme_bw())
figure1 <- addpanel(
  figure = figure1,
  topPanel = 4,
  bottomPanel = 5,
  leftPanel = 1, 
  rightPanel = 3, 
  panel = scramblaseAssayPlot("inputTable.txt") + 
    theme_bw() + 
    xlim(0, 2) + 
    xlab(expression("PPR"~~("mg"/"mmol"))) +
    ylab(expression(P[{phantom()>=1}])))
grid.draw(figure1)
```

Scramblase activity analysis as described above is performed on the data set with
the following calls.

```{r, eval = FALSE}
# Plot PPR plot (Figure 1B)
scramblaseAssayPlot("inputTable.txt") + xlim(0, 2)
# Write out results in tabular manner
scramblaseAssayStats("inputTable.txt")
```

```{r, echo = FALSE}
setwd(tmpDir)
max_PPR <- read.delim(
    "inputTable.txt",
    check.names = FALSE) %>%
  flippant:::calculatePpr() %>%
  max(
    na.rm = TRUE) %>%
  round(
    digits = 2
  )
```

The first call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 1B. Using the customization facilities provided by `ggplot2` 
[@wickham_ggplot2_2009] as a plotting system the plot's x-axis is trimmed to
avoid visual compression of all data by a single data point from the wild type 
rhodopsin series in experiment 1, where $\text{PPR}=`r max_PPR`$. Dashed lines
are added automatically as graphical representations of the fit constant 
$\alpha$, but are difficult to evaluate visually.

Recapitulation of the calculations underlying PPR plot generation by 
`scramblaseAssayPlot` in a call to `scramblaseAssayStats` produces tabular 
output represented in Table 1, which includes series-specific fit constants 
$\alpha$ in numeric and easily accessible form.

```{r, echo = FALSE}
setwd(tmpDir)
ppr_results <- scramblaseAssayStats("inputTable.txt")
ppr_results_alpha_average <- ppr_results %>%
  split(ppr_results$`Experimental Series`) %>% 
  sapply(function(y){mean(y$`Fit Constant`, na.rm = TRUE)}) %>%
  round(2)
ppr_results_alpha_average_names <- names(ppr_results_alpha_average) %>%
  head(n = length(ppr_results_alpha_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(names(ppr_results_alpha_average), n = 1))
ppr_results_alpha_average_print <- ppr_results_alpha_average %>%
  unname() %>%
  head(n = length(ppr_results_alpha_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(ppr_results_alpha_average, n = 1))
knitr::kable(
  ppr_results,
  row.names = FALSE,
  caption = "**Tabular output of *flippant*-based reanalysis of data in Figure 4C/D from @ploier_perturbed_submitted.** An plain text-version of this output is produced by calling `scramblaseAssayStats`.")
```

@@

#Conclusion

```{r, echo = FALSE, cache = TRUE}
setwd(tmpDir)
analysis_timing_n <- 100
analysis_timing_function <- function(){
  scramblaseAssayTraces(
    "inputTable.txt",
    timeMaxSec = 350)
  scramblaseAssayPlot("inputTable.txt")
  scramblaseAssayStats("inputTable.txt")
  gc()
}
analysis_timing <- microbenchmark(
  analysis_timing_function(),
  times = analysis_timing_n)

@@

analysis_timing_iqr <- round(
  diff(
    unlist(
      summary(analysis_timing)[c("lq", "uq")],
      use.names = FALSE)),
  2)
```

With *flippant* we present an integrated solution for data analysis in the 
context of dithionite scramblase assays. Requiring only basic familiarity with 
R as an environment for statistical computing and graphics 
[@r_core_team_r:_2015], scientists can quickly analyse such data and arrive at 
publication-grade graphics that offer extensive facilities for individual 
optimization and adaptation [@wickham_ggplot2_2009]. On a laptop system with 
`r parallel::detectCores()` cores and 8 GB RAM (no SSD), running `Linux` and 
R version `r paste(R.version[c("major","minor")], collapse = ".")`, the
statements generating equivalents to the plots and calculations presented here
take a mere `r analysis_timing_median` seconds (median; 
$\text{IQR}=`r analysis_timing_iqr`~\text{s}$; $n=`r analysis_timing_n`$).
